Evaluating the Decency and Consistency of Data Validation Tests Generated by LLMs

### 2023-09-29 paper/paper.qmd
lindsay edits

The Investigative Journalism Foundation (IJF) created and maintains a dataset related to political donations in Canada. As of September 2023, the dataset comprises 9,204,112 observations and 14 variables. Every day new observations are added, based on newly released donations records in the past time period that are made available by the provincial and federal elections agencies. This release is variable and a source may not post records for a few weeks and then many are made available. @katzandmoore manually construct an extensive suite of automated tests for this dataset. These impose certain minimum standards on the dataset, including: that constituent aspects add to match any total, class is appropriate, and null values are where expected. This suite allows researchers to use the dataset with confidence and ensures that new additions are fit for purpose.

We revisit this suite of tests to determine whether we can use Large Language Models (LLMs) to mimic this suite of validation tests. We consider a variety of prompts, roles, learning and temperature settings, resulting in 96 total experiments. In particular, we consider four prompt variations: ask for expectations; ask for expectations given described context; ask for expectations having first asked for simulation; ask for expectations giving a sample of data. We also consider zero-, one-, and few-shot learning; four temperature values: 0, 0.4, 0.6, and 1; and two roles: helpful assistant, and expert data scientist. For every combination we obtain five responses from the LLM. We run all this separately for both GPT-3.5 and GPT-4.

@@

We find there are considerable returns to few-shot learning, and that the more explicit the data setting can be the better. **[Add more]**

@@

**Schwartz - can you please add one paragraph of background about the IJF and why high-quality datasets are important for Canadian democracy/journalism/whatever.**

Canadian legislation requires political parties and candidates to disclose records of financial contributions they receive. These records are maintained by elections agencies across provinces and territories, and at the Federal level. The frequency and scope of these disclosures vary across jurisdictions. For instance, in British Columbia, parties, candidates, constituency associations and leadership and nomination contestants can receive political donations, while in Newfoundland and Labrador, donation recipients are limited to only parties and candidates (IJF, 2023).

The IJF’s political donations database is a compilation of these political financing records across all Canadian jurisdictions, with data spanning from 1993 to the present day. The database contains 14 variables including the donor’s name, the political party and entity to which the donation was made, the amount donated, as well as the region and year of the donation.

While the IJF’s database is available in a clean, user-friendly format, the original records upon which it was created were not all accessible in this way. The format of donation records varies across jurisdiction and time. While some are available in readily downloadable spreadsheets, others are available as PDF and HTML files — the former necessitating the use of optical character recognition (OCR) technology (IJF, 2023). To prepare their database for publication, the IJF team performed significant manual cleaning. The majority of this work resulted from the conversion of PDF donation records to rectangular CSV format using OCR which is prone to scanning-related errors, such as the number 0 being scanned as the letter O. The IJF manually corrected these errors wherever they were identified by carefully comparing the machine-legible OCR output to the original static PDF donations record (IJF, 2023).

Additional cleaning was done for the purpose of data legibility. For instance, the IJF standardized donation dates to match the YYYY-MM-DD format, and they standardized donor names which were in the format “Surname, first name” to be in the form of “First name surname” (IJF, 2023). Party names and donor types were also standardized for consistency, and donation records with an abbreviated party name were supplied with the complete name for that party. Finally, in rows where the donor type was null but only individuals were legally allowed to make donations in that jurisdiction and year, the IJF changed that null entry to be “Individual” (IJF, 2023).

Despite the thoughtful and thorough cleaning performed by the IJF team, the magnitude of these data coupled with their self-reported nature makes them prone to both human error, and computational error arising from the parsing process.

the use of computational tools to assess data quality.

With over 9.2 million rows in this database, it would be a massive undertaking for the IJF to manually

**Katz - You've already added the paragraphs above, but please add whatever more you think relevant.**

@@

-  Four prompts:
````
    - The Investigative Journalism Foundation (IJF) created and maintains a CSV dataset related to political donations in Canada. Each observation in the dataset is a donation, and the dataset has the following variables: "index", "amount", "donor_location", "donation_date", "donor_full_name", "donor_type", "political_entity", "political_party", "recipient", "region", "donation_year", "amount_monetary", "amount_non_monetary", "electoral_event", "electoral_district", "added".\n\rPlease write a series of expectations using the Python package great_expectations for this dataset.
    - The Investigative Journalism Foundation (IJF) created and maintains a CSV dataset related to political donations in Canada. Each observation in the dataset is a donation, and the dataset has the following variables: "index", "amount", "donor_location", "donation_date", "donor_full_name", "donor_type", "political_entity", "political_party", "recipient", "region", "donation_year", "amount_monetary", "amount_non_monetary", "electoral_event", "electoral_district", "added". 
        - "amount" is a monetary value that cannot be less than $0. An example observation is "195.46". It is, possible, but unlikely to be more than $1,000.00. It cannot be NA. It should be a numeric. The maximum donation "amount" depends on the value of "region" and "year". For "Federal" is 1675, for "Quebec" is 100 since 2013 and 500 for earlier years, for "British Columbia" is 1309.09, for "Ontario" is 3325, and for "Alberta" is 4300. There is no limit for "Saskatchewan".
        - "amount" should be equal to the sum of "amount_monetary" and "amount_non_monetary".
        - "region" can be one of the following values: "Federal", "Quebec", "British Columbia", "Ontario", "Saskatchewan", "Alberta". It cannot be NA. It should be a factor variable.
        - "donor_full_name" is a string. It cannot be NA. It is usually a first and last name, but might also include a middle initial. It should be in title case.
        - "donation_date" should be a date in the following format: YYYY-MM-DD. It could be NA. The earliest donation is from 2010-01-01. The latest donation is from 2023-09-01.
        - "donation_year" should match the year of "donation_date" if "donation_date" is not NA, but it is possible that "donation_year" exists even if "donation_date" does not. The earliest year is 2010 and the latest year is 2023. This variable is an integer.
        - "political_party" cannot be NA. It should be a factor that is equal to one of: "New Democratic Party", "Liberal Party of Canada", "Conservative Party of Canada".

@@

        Based on this sample please write a series of expectations using the Python package great_expectations for this dataset.
````
- Zero-, one-, and few-shot learning:
````
    - The following text in quotes is an example of an expectation for this dataset:
        """
        # Check that there is nothing null in any column of donations details
        donations_mv.expect_column_values_to_not_be_null(column='donor_full_name')
        """
    - The following text in quotes is an example of three expectations for this dataset:
        """
        # Check that there is nothing null in any column of donations details
        donations_mv.expect_column_values_to_not_be_null(column='donor_full_name')
        # Check that the federal donation does not exceed the maximum
        donations_mv.expect_column_values_to_be_between(
            column = 'amount',
            max_value = 1675,
            row_condition = 'region=="Federal" & donor_full_name.str.contains("Contributions Of")==False & donor_full_name.str.contains("Estate Of")==False & donor_full_name.str.contains("Total Anonymous Contributions")==False & donation_year == 2022 & political_entity.str.contains("Leadership")==False',
            condition_parser = 'pandas'
        )
        # Check that the date matches an appropriate regex format
        donations_mv.expect_column_values_to_match_regex(column = 'donation_date',
                                                    regex = '\\d{4}-\\d{2}-\\d{2}',
                                                    row_condition = "donation_date.isna()==False",
                                                    condition_parser = 'pandas')
        """
````
- Four temperature values: 0, 0.4, 0.6, and 1. 
- Two roles: 
````
    - You are a helpful assistant.
    - You are a highly-trained, experienced, data scientist who is an expert at writing readable, correct, Python code.
````

@@

This results in a dataset of responses. The option that gave rise to each response was blinded, and the order randomized, and then they were ranked by one experienced human coder on two metrics. The first, "consistency", was a ranking 1-5, of how different each of the five responses was for that particular combination of variables. 1 means that Responses 1-5 were wildly different. 5 means that Responses 1-5 are entirely or essentially the same. The human coder then ranked the "decency" of the first response for each combination of variables. This is a measure of how different the LLM responses were, compared with the code written by the experienced data scientist who wrote the original suite of tests. The LLM responses are not expected to have the full context of the code, so we do not expect an exact match, but it should actually write code, import relevant libraries, add comments, deal with class, write a variety of relevant expectations. 1 means that the code is unusable, 2 means that it is not unusable but would need a lot of work and would be disappointing from a human, 3 means that it is fine but would need some fixing, 4 means it is broadly equivalent to what the gold standard suite, and 5 means it is in no worse and is better in some way than the gold standard validation suite.

@@

@fig-version examines how decency and consistency differ based on whether GPT-3.5 or GPT-4 is used. Unexpectedly, GPT-4 has fewer responses rated 5/5, compared with GPT-3.5 (@fig-version-1). GPT-3.5 has fewer responses rated 2/5, but overall the mean decency response for GPT-3.5 is 3.23, while the mean decency of the responses generated by GPT-4 is 3.01. The consistency is not too different between the two versions, with GPT-3.5 having an average of 3.61, while GPT-4 has an average of 3.65. GPT-4 appears to have fewer responses that are completely identical (@fig-version-2).

@@

@fig-prompt examines how decency and consistency differ based on which prompt is used. The prompts differ by how much information is provided. The least informative prompt, "Name", essentially consists of just providing the LLM with thee names of the columns that are expected. The next most informative prompt, "Describe", adds a detailed description of what we expect of the observations. "Simulate" adds that we expect the LLM to first simulate a dataset based on that description, before generating the expectations. And finally, the most informative prompt, "Example", provides a snapshot of the dataset, consisteing of the relevant variables and ten observations.

@@

## Why validation tests are important and some of the challenges of writing them

**Katz - can you please add a page or two that interacts with the literature and situates this contribution? And whatever else you think. Feel free to change the sub-heading**

>>>>

The Investigative Journalism Foundation (IJF) created and maintains a dataset related to political donations in Canada. As of September 2023, the dataset comprises 9,204,112 observations and 15 variables. Every day new observations are added, based on newly released donations records in the past time period that are made available by the provincial and federal elections agencies. This release is variable and a source may not post records for a few weeks and then many are made available. @katzandmoore manually construct an extensive suite of automated tests for this dataset. These impose certain minimum standards on the dataset, including: that constituent aspects add to match any total, class is appropriate, and null values are where expected. This suite allows researchers to use the dataset with confidence and ensures that new additions are fit for purpose.

We revisit this suite of tests to determine whether we can use Large Language Models (LLMs) to mimic this suite of validation tests. We consider a variety of prompts, roles, learning and temperature settings, resulting in 96 total experiments. In particular, we consider four prompt variations: ask for expectations; ask for expectations given described context; ask for expectations having first asked for a simulation; ask for expectations given a sample of data. We also consider zero-, one-, and few-shot learning; four temperature values: 0, 0.4, 0.6, and 1; and two roles: helpful assistant, and expert data scientist. For every combination we obtain five responses from the LLM. We run all experiments separately for both GPT-3.5 and GPT-4.

@@

We find there are considerable returns to few-shot learning, and that the more explicit the data setting can be the better. **\[Add more\]**

@@

The Investigative Journalism Foundation (IJF) is a nonprofit news media outlet that is centered around public interest journalism. Their mandate is to help rebuild trust in Canadian democracy and hold the powerful accountable through data-driven investigative reporting. A core component of the IJF's work is building and actively maintaining eight publicly available databases with information on political donations, registered charities, and political lobbying in Canada. While the information that these databases are built upon are ostensibly public, they are not maintained or available in a way that is widely accessible or conducive to analysis. Further, the format and accessibility of these data vary greatly over time and across jurisdictions, making it incredibly difficult to look at temporal or regional differences in the data. In turn, the IJFs collation of these databases and high-quality journalism informed by these data serves as a crucial contribution to rebuilding trust and transparency in Canadian democracy.

One of the IJFs eight databases, and the focus of this work, is the political donations database. Canadian legislation requires political parties and candidates to disclose records of financial contributions they receive. These records are maintained by elections agencies across provinces and territories, and at the Federal level. The frequency and scope of these disclosures vary across jurisdictions. For instance, in British Columbia, parties, candidates, constituency associations and leadership and nomination contestants can receive political donations, while in Newfoundland and Labrador, donation recipients are limited to only parties and candidates [@ijfmethods]. The IJF's political donations database is a compilation of these political financing records across all Canadian jurisdictions, with data spanning from 1993 to the present day. The database contains 14 variables including the donor's name, the political party and entity to which the donation was made, the amount donated, as well as the region and year of the donation.

While the IJF's database is available in a clean, user-friendly format, the original records upon which it was created were not all accessible in this way. The format of donation records varies across jurisdiction and time. While some are available in readily downloadable spreadsheets, others are available as PDF and HTML files --- the former necessitating the use of optical character recognition (OCR) technology [@ijfmethods]. To prepare their database for publication, the IJF team performed significant manual cleaning. The majority of this work resulted from the conversion of PDF donation records to rectangular CSV format using OCR which is prone to scanning-related errors, such as the number 0 being scanned as the letter O. The IJF manually corrected these errors wherever they were identified by carefully comparing the machine-legible OCR output to the original static PDF donations record [@ijfmethods].

Additional cleaning was done for the purpose of data legibility. For instance, the IJF standardized donation dates to match the YYYY-MM-DD format, and they standardized donor names which were in the format "Surname, first name" to be in the form of "First name surname" [@ijfmethods]. Party names and donor types were also standardized for consistency, and donation records with an abbreviated party name were supplied with the complete name for that party. Finally, in rows where the donor type was null but only individuals were legally allowed to make donations in that jurisdiction and year, the IJF changed that null entry to be "Individual" [@ijfmethods].

Despite the thoughtful and thorough cleaning performed by the IJF team, the magnitude of these data coupled with their self-reported nature makes them prone to both human error and computational error arising from the parsing process. With over 9.2 million rows in this database, it would be a massive undertaking for the IJF to manually identify all errors and inconsistencies present in their data - whether that is as minor as an incorrectly formatted date, or as major as a reported donation amount thousands of dollars above the legal limit. To address this challenge, the IJF has been exploring the use of computational tools to assess data quality [@katzandmoore]. The team has developed a suite of comprehensive data tests for all eight databases using Python's [Great Expectations](https://greatexpectations.io/) platform, which contains a number of pre-defined functions to test that particular characteristics expected of a dataset hold true in the data at large.

As described by @katzandmoore, the process of developing this test suite was iterative in nature, and significant domain knowledge was necessary to develop tests which were both accurate and valuable. For the political donations database, @katzandmoore developed a comprehensive suite of tests pertaining to missingness, formatting, and value expectations in the data. For instance, they set the expectation that for donations made in British Columbia, Ontario, or Federally, the `donation_date` entry should not be missing, because those are the only three jurisdictions which collect that variable [@katzandmoore]. They also test that donations data from 2022 align with the 2022 legal donation limits for each jurisdiction, and that, where applicable, the sum of reported monetary and non-monetary contribution amounts add up to the total reported donation amount. The [article](https://arxiv.org/abs/2309.12886) written by @katzandmoore provides complete details on the development and implementation of these tests.

@@

-   Four prompts:

```{=html}
<!-- -->
```
        - The Investigative Journalism Foundation (IJF) created and maintains a CSV dataset related to political donations in Canada. Each observation in the dataset is a donation, and the dataset has the following variables: "index", "amount", "donor_location", "donation_date", "donor_full_name", "donor_type", "political_entity", "political_party", "recipient", "region", "donation_year", "amount_monetary", "amount_non_monetary", "electoral_event", "electoral_district", "added".\n\rPlease write a series of expectations using the Python package great_expectations for this dataset.
        - The Investigative Journalism Foundation (IJF) created and maintains a CSV dataset related to political donations in Canada. Each observation in the dataset is a donation, and the dataset has the following variables: "index", "amount", "donor_location", "donation_date", "donor_full_name", "donor_type", "political_entity", "political_party", "recipient", "region", "donation_year", "amount_monetary", "amount_non_monetary", "electoral_event", "electoral_district", "added". 
            - "amount" is a monetary value that cannot be less than $0. An example observation is "195.46". It is, possible, but unlikely to be more than $1,000.00. It cannot be NA. It should be a numeric. The maximum donation "amount" depends on the value of "region" and "year". For "Federal" is 1675, for "Quebec" is 100 since 2013 and 500 for earlier years, for "British Columbia" is 1309.09, for "Ontario" is 3325, and for "Alberta" is 4300. There is no limit for "Saskatchewan".
            - "amount" should be equal to the sum of "amount_monetary" and "amount_non_monetary".
            - "region" can be one of the following values: "Federal", "Quebec", "British Columbia", "Ontario", "Saskatchewan", "Alberta". It cannot be NA. It should be a factor variable.
            - "donor_full_name" is a string. It cannot be NA. It is usually a first and last name, but might also include a middle initial. It should be in title case.
            - "donation_date" should be a date in the following format: YYYY-MM-DD. It could be NA. The earliest donation is from 2010-01-01. The latest donation is from 2023-09-01.
            - "donation_year" should match the year of "donation_date" if "donation_date" is not NA, but it is possible that "donation_year" exists even if "donation_date" does not. The earliest year is 2010 and the latest year is 2023. This variable is an integer.
            - "political_party" cannot be NA. It should be a factor that is equal to one of: "New Democratic Party", "Liberal Party of Canada", "Conservative Party of Canada".

@@

            Based on this sample please write a series of expectations using the Python package great_expectations for this dataset.

-   Zero-, one-, and few-shot learning:

```{=html}
<!-- -->
```
        - The following text in quotes is an example of an expectation for this dataset:
            """
            # Check that there is nothing null in any column of donations details
            donations_mv.expect_column_values_to_not_be_null(column='donor_full_name')
            """
        - The following text in quotes is an example of three expectations for this dataset:
            """
            # Check that there is nothing null in any column of donations details
            donations_mv.expect_column_values_to_not_be_null(column='donor_full_name')
            # Check that the federal donation does not exceed the maximum
            donations_mv.expect_column_values_to_be_between(
                column = 'amount',
                max_value = 1675,
                row_condition = 'region=="Federal" & donor_full_name.str.contains("Contributions Of")==False & donor_full_name.str.contains("Estate Of")==False & donor_full_name.str.contains("Total Anonymous Contributions")==False & donation_year == 2022 & political_entity.str.contains("Leadership")==False',
                condition_parser = 'pandas'
            )
            # Check that the date matches an appropriate regex format
            donations_mv.expect_column_values_to_match_regex(column = 'donation_date',
                                                        regex = '\\d{4}-\\d{2}-\\d{2}',
                                                        row_condition = "donation_date.isna()==False",
                                                        condition_parser = 'pandas')
            """

-   Four temperature values: 0, 0.4, 0.6, and 1.
-   Two roles:

```{=html}
<!-- -->
```
        - You are a helpful assistant.
        - You are a highly-trained, experienced, data scientist who is an expert at writing readable, correct, Python code.

@@

This results in a dataset of responses. The option that gave rise to each response was blinded, and the order randomized, and then they were ranked by one experienced human coder on two metrics. The first, "consistency", was a ranking 1-5, of how different each of the five responses was for that particular combination of variables. 1 means that responses 1-5 were wildly different. 5 means that responses 1-5 are entirely or essentially the same. The human coder then ranked the "decency" of the first response for each combination of variables. This is a measure of how different the LLM responses were, compared with the code written by the experienced data scientist who wrote the original suite of tests. The LLM responses are not expected to have the full context of the code, so we do not expect an exact match, but it should actually write code, import relevant libraries, add comments, deal with class, and write a variety of relevant expectations. 1 means that the code is unusable, 2 means that it is not unusable but would need a lot of work and would be disappointing from a human, 3 means that it is fine but would need some fixing, 4 means it is broadly equivalent to what the gold standard suite contains, and 5 means it is in no worse and is better in some way than the gold standard validation suite.

@@

@fig-version examines how decency and consistency differ based on whether GPT-3.5 or GPT-4 is used. Unexpectedly, GPT-4 has fewer responses rated 5/5 for decency, compared with GPT-3.5 (@fig-version-1). GPT-3.5 has fewer responses rated 2/5, but overall the mean decency response for GPT-3.5 is 3.23, while the mean decency of the responses generated by GPT-4 is 3.01. The consistency is not too different between the two versions, with GPT-3.5 having an average of 3.61, while GPT-4 has an average of 3.65. GPT-4 appears to have fewer responses that are completely identical (@fig-version-2).

@@

@fig-prompt examines how decency and consistency differ based on which prompt is used. The prompts differ by how much information is provided. The least informative prompt, "Name", essentially consists of just providing the LLM with the names of the columns that are expected to be in the dataset. The next most informative prompt, "Describe", adds a detailed description of what we expect of the observations. "Simulate" adds that we expect the LLM to first simulate a dataset based on that description, before generating the expectations. And finally, the most informative prompt, "Example", provides a snapshot of the dataset, consisting of the relevant variables and ten observations.

@@

## On the importance and challenges of writing validation tests
**Katz - can you please add a page or two that interacts with the literature and situates this contribution? And whatever else you think. Feel free to change the sub-heading**

Much important work has been done to illustrate the importance of reproducibility and reliability in data science across domains.

importance - promotes transparency
           - promotes trustworthiness and credibility
           - promotes a more thorough understanding of the data

The value of data validation is not limited to particular domains --- it is a fundamental component of a reliable data science workflow, regardless of the data at hand.

There are a number of challenges that accompany the development and implementation of data validation tests. First, implementing automated tests requires the ability to develop code [@zhang20]. While there are a number of accessible, well-documented testing-centered libraries in different programming languages (e.g., [Great Expectations](greatexpectations.io), [pointblank](https://cran.r-project.org/web/packages/pointblank/index.html)), the implementation of a test suite may require significant data processing work. One of the main takeaways from @katzandmoore on their data validation efforts for the IJF is that data tests should not be compromised just to be compatible with pre-defined test functions that exist in a software package. In order to develop the most valuable suite of tests, it may be necessary to create new variables in a dataset or to merge multiple datasets, for example. This presents a challenge for individuals who are not trained in or comfortable with computer programming, but wish to programmatically test their data. In addition, the development of a comprehensive, bespoke, and accurate data test suite can rarely be done by one data scientist alone. As advocated by @katzandmoore, domain expertise is fundamental for the creation and development of high-quality data tests. For instance, without incorporating the knowledge that donation date is only collected in three jurisdictions (British Columbia, Ontario and Federal) into their code, @katzandmoore's test surrounding missingness for that variable would have produced inflated failure rates. This is knowledge that was gained through collaborative efforts with IJF team members who were involved in dataset construction. For an individual working on an unfamiliar dataset, seeking out domain knowledge and gaining a thorough understanding of ones data is a non-trivial task in and of itself. Finally, the process of designing, developing, and deploying data validation tests is iterative in nature and takes a great deal of time. For researchers or journalists who are eager to publish their findings for instance, there may not be enough incentive or resources for them to want to do this time-consuming validation work. This challenge is only amplified if an individual was not involved in the dataset construction process and does not have access to vital domain expertise, or if an individual does not have programming experience.

Evidently, while data validation is incredibly valuable for the production of transparent and trustworthy research, it comes with many challenges which hinder its accessibility. This work presents an opportunity to promote data validation through the use of LLMs. While there has been a great deal of work done to develop more accessible programming tools for data validation, these tools do not entirely mitigate the challenges which accompany data validation, namely the need for programming knowledge, domain expertise, and a great deal of time. Our findings illustrate the ability of LLMs to rapidly produce **[valuable? usable? elementary?]** data validation tests, written in code, with limited information.
